| Base de datos de abejas ibéricas \n| Iberian Bees database\n

### 2022-04-07 Manuscript/manuscript.Rmd
Edits

  chunk_output_type: console
header-includes:
- \DeclareUnicodeCharacter{0301}{*************************************}
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \captionsetup[table]{labelformat=empty}
---

@@

> **Iberian Bees database**. Bees are a diverse group with more than 1,000 species known from the Iberian Peninsula. They have increasingly received special attention due to their important role as pollinators and providers of ecosystem services. In addition, various rapid human-induced environmental changes are leading to the decline of some of its populations. However, we know very little about the conservation status of most species and for many species, we hardly know their true distributions across the Iberian Peninsula. Here we present a collaborative effort to collate and curate a database of Iberian bee occurrences to answer questions about their distribution, habitat preference, phenology, or historical trends. In total we have accumulated `r formatC(nrow(data),format="d", big.mark=",")` records from the Iberian Peninsula and the Balearic Islands of `r length(unique(data$Accepted_name))` different species with `r round(geo/nrow(data),2)*100`% of georeferenced records collected between `r min(data$Year, na.rm = T)` and `r max(data$Year, na.rm = T)`. In addition, each record has associated information such as the sampling location (`r round(local/nrow(data),2)*100`%), collector and person who id the species (`r round(deter/nrow(data),2)*100`%), date of the capture (`r round(date/nrow(data),2)*100`%) and plant species where the bees was captured (`r round(flw/nrow(data),2)*100`%). We believe that this database is the starting point to better understand and conserve bee biodiversity in the Iberian Peninsula. It can be accessed at: https://doi.org/10.5281/zenodo.6354502.

@@

> Biodiversidad; Apoidea; Polinizadores; Fenología

@@

> Biodiversity; Apoidea; Pollinators; Phenology

@@

Bees (Hymenoptera, Apoidea, Anthophila) are a diverse group of insects with more than 20,000 species described worldwide [@ascher2020]. The Iberian Peninsula, with its bee-loved Mediterranean climate, is one of the hotspots of bee biodiversity with more than 1,000 species recorded to date and counting [@ascher2020; @ortiz2020]. Unfortunately, despite the high biodiversity of Iberian bees, we know very little about the species themselves and their distributions. This is paradoxical, as in recent years bees have been highlighted as a keystone group in the ecosystem due to a key function they provide, the pollination of thousands of plants [@ollerton2011], including most crop species [@klein2007]. In addition, various rapid human-induced environmental changes are leading to the decline of some bee populations [@goulson2015]. However, the response of bee species to global change is heterogeneous both in space and also taxonomically. That is, while in some areas species are declining drastically  [@burkle2013], other areas are well conserved [@herrera2019] and while some species are in the brink of extinction [@cameron2011] some species are even thriving [@russo2016]. A better description of bee distributions, habitat preferences, phenologies, or historical trends in the Iberian Peninsula would require information on when and where different species occur.

@@

Here we have compiled an Iberian Bee Database of occurrence records by integrating digitalization efforts by leading national history museums, with more than 100 individual datasets contributed by researchers, and publicly available information on online repositories. After data cleaning and harmonization, we release the first `r nrow(data)` records from the Iberian Peninsula and the Balearic Islands. **Figure 1** shows the spatial distribution of records, which as expected its biased to certain areas near main cities. Overall, the database contains information of `r length(unique(data$Accepted_name))` different species, for which the most commonly collected are depicted in **Figure 2**. The data has been collected between `r min(data$Year, na.rm = T)` and `r max(data$Year, na.rm = T)`, with an increasing trend of the number of records with time, reflecting a renewed interest for documenting this taxa (**Figure 3A**). As expected, most records were recorded in spring with May being the month with highest number of records (**Figure 3B**). More than `r round(geo/nrow(data),2)*100`% of these records have associated information such as the location and date of capture, and `r round(flw/nrow(data),2)*100`% have extra data such as the plant species where it was collected. Here, we release the version 1.0 of the Iberian Bee Database with name "IberianBees". Although it will continue to grow, we believe that this is a good starting point to better understand and conserve bee biodiversity. It can be accessed at: https://doi.org/10.5281/zenodo.6354502.

@@

Las abejas (Hymenoptera, Apoidea) son un grupo muy diverso de especies con más de 20,000 especies descritas en el mundo [@ascher2020]. Debido a las condiciones climáticas de la península ibérica, esta se caracteriza por ser uno de los puntos calientes de biodiversidad a nivel mundial con más de 1000 abejas descritas hasta la fecha [@ortiz2020]. A pesar de la gran biodiversidad de abejas ibéricas, conocemos muy poco de su ecología y distribución. Paradójicamente las abejas son un un grupo clave que desarrolla una función vital para la moyoría de ecosistemaas terrestres, la polinización de miles de plantas [@ollerton2011] incluyendo a la mayoría de cultivos [@klein2007]. Además el cambio debido al impacto del ser humano esta dando lugar al declive de algunas de sus poblaciones [@goulson2015]. No obstante, la respuesta de las abejas al cambio global es heterogénea tanto en el espacio como taxonomicámente. Es decir, existen áreas que están sufriendo un decaimiento poblacional drástico [@burkle2013] mientras que otras se mantienen bien conservadas [@herrera2019] y mientras que algunas especies se encuntra al borde de la extinción [@cameron2011], otras se encuentran incluso prosperando [@russo2016]. Una mejor descripción de la distribución de las abejas sobre su distribución, preferencia de hábitat, fenología, o tendencias históricas en la península ibérica requerirían información de dónde y cuándo las especies aparecen.

@@

```{r write_citations, cache=FALSE, include=FALSE}
write.bibtex(file="knitcitations.bib")
```

@@

**Figura 1**. Barplot of the number of records of the 20 most common species.

**Figura 2**. Map of the Iberian Peninsula where the different number of records are shown with a colour code in logarithmic scale. Areas with high number of records are coloured in yellow, areas with low number of records in purple and areas with lack of records are shown in white.

@@

**Figura 1**. Gráfico de barras del número de registros de las 20 especies más abundantes.

**Figura 2**. Mapa de la península ibérica donde se muestra el número de registros en escala logarítmica con un código de colores. Áreas con alto número de registros se muestran en amarillo, con un bajo número de registros en morado y zonas con ausencia de registros en blanco.

@@

```{r echo=FALSE, fig.align='center', out.width="70%", fig.cap="\\textbf{Figure 1}"}
# Prepare top 20 records
top_20 <- data %>%
  filter(!Accepted_name == "Apis mellifera") %>%
  group_by(Accepted_name) %>%
  summarise(no_rows = length(Accepted_name)) %>%
  arrange(-no_rows) %>%
  slice(1:20)
# Fix row position
top_20$Accepted_name <- factor(top_20$Accepted_name, levels = top_20$Accepted_name)
# Plot
ggplot(top_20, aes(x = Accepted_name, y = no_rows)) +
  geom_bar(stat = "identity", fill = "#287D8EFF") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8, face = "italic")) +
  ylab("Number of records") +
  xlab("Species") +
  scale_y_continuous(expand = c(0, 0))
```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap="\\textbf{Figure 2}",out.width="70%"}
# code courtesy of Paco Rodriguez.
library(sf)
library(raster)
library(rasterVis)
library(tidyverse)
crs.geo <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
crs.laea <- "+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"

@@

>>>>

  chunk_output_type: console
header-includes:
- \DeclareUnicodeCharacter{0301}{*************************************}
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
- \captionsetup[table]{labelformat=empty}

---

@@

> **Iberian Bees database**. Bees are a diverse group with more than 1,000 species known from the Iberian Peninsula. They have increasingly received special attention due to their important role as pollinators and providers of ecosystem services. In addition, various rapid human-induced environmental changes are leading to the decline of some of its populations. However, we know very little about the conservation status of most species and for many species, we hardly know their true distributions across the Iberian Peninsula. Here we present a collaborative effort to collate and curate a database of Iberian bee occurrences to answer questions about their distribution, habitat preference, phenology, or historical trends. In total we have accumulated `r formatC(nrow(data),format="d", big.mark=",")` records from the Iberian Peninsula and the Balearic Islands of `r length(unique(data$Accepted_name))` different species with `r round(geo/nrow(data),2)*100`% of georeferenced records collected between `r min(data$Year, na.rm = T)` and `r max(data$Year, na.rm = T)`. In addition, each record has associated information such as the sampling location (`r round(local/nrow(data),2)*100`%), collector and person who identified the species (`r round(deter/nrow(data),2)*100`%), date of the capture (`r round(date/nrow(data),2)*100`%) and plant species where the bees were captured (`r round(flw/nrow(data),2)*100`%). We believe that this database is the starting point to better understand and conserve bee biodiversity in the Iberian Peninsula. It can be accessed at: https://doi.org/10.5281/zenodo.6354502.

@@

> Biodiversidad; Apoidea; Anthophila; Polinizadores; Fenología

@@

> Biodiversity; Apoidea; Anthophila; Pollinators; Phenology

@@

Bees (Hymenoptera, Apoidea, Anthophila) are a diverse group of insects with more than 20,000 species described worldwide (Ascher & Pickering, 2020. The Iberian Peninsula, with its bee-loved Mediterranean climate, is one of the hotspots of bee biodiversity with more than 1,000 species recorded to date and counting  (Ascher & Pickering, 2020; Ortiz-Sánchez, 202). Unfortunately, despite the high biodiversity of Iberian bees, we know very little about the species themselves and their distributions. This is paradoxical, as in recent years bees have been highlighted as a keystone group in the ecosystem due to a key function they provide, the pollination of thousands of plants [@ollerton2011], including most crop species [@klein2007]. In addition, various rapid human-induced environmental changes are leading to the decline of some bee populations [@goulson2015]. However, the response of bee species to global change is heterogeneous both in space and also taxonomically. That is, while in some areas species are declining drastically  [@burkle2013], other areas are well conserved [@herrera2019] and while some species are in the brink of extinction [@cameron2011] some species are even thriving [@russo2016]. A better description of bee distributions, habitat preferences, phenologies, or historical trends in the Iberian Peninsula would require information on when and where different species occur.

@@

Here we have compiled an Iberian Bee Database of occurrence records by integrating digitalization efforts by leading national history museums, with more than 100 individual datasets contributed by researchers, and publicly available information on online repositories. After data cleaning and harmonization, we release the first `r formatC(nrow(data),format="d", big.mark=",")` records from the Iberian Peninsula and the Balearic Islands. **Figure 1** shows the spatial distribution of records, which as expected its biased to certain areas near main cities. Overall, the database contains information of `r length(unique(data$Accepted_name))` different species, for which the most commonly collected are depicted in **Figure 2**. The data has been collected between `r min(data$Year, na.rm = T)` and `r max(data$Year, na.rm = T)`, with an increasing trend of the number of records with time, reflecting a renewed interest for documenting this taxa (**Figure 3A**). As expected, most records were recorded in spring with May being the month with highest number of records (**Figure 3B**). More than `r round(geo/nrow(data),2)*100`% of these records have associated information such as the location and date of capture, and `r round(flw/nrow(data),2)*100`% have extra data such as the plant species where bee specimens were collected. Here, we release the version 1.0 of the Iberian Bee Database with name "IberianBees". Although it will continue to grow, we believe that this is a good starting point to better understand and conserve bee biodiversity. It can be accessed at: https://doi.org/10.5281/zenodo.6354502.

@@

Las abejas (Hymenoptera, Apoidea) son un grupo muy diverso de especies con más de 20,000 especies descritas en el mundo [@ascher2020]. Debido a las condiciones climáticas de la península ibérica, esta se caracteriza por ser uno de los puntos calientes de biodiversidad a nivel mundial con más de 1000 abejas citadas hasta la fecha [@ortiz2020]. A pesar de la gran biodiversidad de abejas ibéricas, conocemos muy poco de su ecología y distribución. Paradójicamente las abejas son un grupo clave que desarrolla una función vital para la mayoría de ecosistemaas terrestres, la polinización de miles de plantas [@ollerton2011] incluyendo a la mayoría de cultivos [@klein2007]. Además el cambio debido al impacto del ser humano esta dando lugar al declive de algunas de sus poblaciones [@goulson2015]. No obstante, la respuesta de las abejas al cambio global es heterogénea tanto en el espacio como taxonomicámente. Es decir, existen áreas que están sufriendo un decaimiento poblacional drástico [@burkle2013] mientras que otras se mantienen bien conservadas [@herrera2019] y mientras que algunas especies se encuentran al borde de la extinción [@cameron2011], otras se encuentran incluso prosperando [@russo2016]. Una mejor descripción de la distribución de las abejas sobre su distribución, preferencia de hábitat, fenología, o tendencias históricas en la península ibérica requerirían información de dónde y cuándo las especies aparecen.

@@

@@

**Figura 1**. Map of the Iberian Peninsula where the different number of records are shown with a colour code in logarithmic scale. Areas with high number of records are coloured in yellow, areas with low number of records in purple and areas with lack of records are shown in white.

**Figura 2**. Barplot of the number of records of the 20 most common species.

@@

**Figura 1**. Mapa de la península ibérica donde se muestra el número de registros en escala logarítmica con un código de colores. Áreas con alto número de registros se muestran en amarillo, con un bajo número de registros en morado y zonas con ausencia de registros en blanco.

**Figura 2**. Gráfico de barras del número de registros de las 20 especies más abundantes.

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.cap="\\textbf{Figure 1}",out.width="70%"}
# code courtesy of Paco Rodriguez.
library(sf)
library(raster)
library(rasterVis)
library(tidyverse)
crs.geo <- "+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs"
crs.laea <- "+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs"

@@

```{r echo=FALSE, fig.align='center', out.width="70%", fig.cap="\\textbf{Figure 2}"}
# Prepare top 20 records
top_20 <- data %>%
  filter(!Accepted_name == "Apis mellifera") %>%
  group_by(Accepted_name) %>%
  summarise(no_rows = length(Accepted_name)) %>%
  arrange(-no_rows) %>%
  slice(1:20)
# Fix row position
top_20$Accepted_name <- factor(top_20$Accepted_name, levels = top_20$Accepted_name)
# Plot
ggplot(top_20, aes(x = Accepted_name, y = no_rows)) +
  geom_bar(stat = "identity", fill = "#287D8EFF") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8, face = "italic")) +
  ylab("Number of records") +
  xlab("Species") +
  scale_y_continuous(expand = c(0, 0))
```
